Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels

### 2021-08-06 manuscript/manuscript.Rmd
update

---
title: Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
    footnote: 
  - name: Alan Robley
    email: 
    affiliation: ARI
  - name: Nick Golding
    email: 
    affiliation: CU,TKI
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
  - code: ARI 
    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
  - code: CU 
    address: Curtin University, Bentley, WA, Australia
  - code: TKI 
    address: Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  1. Background.
  2. Methods.
  3. Results.
  4. *Synthesis and applications.*

@@

Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes. These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time. This is key because species interactions and niche preferences are multidimensional.

Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g. fire). Animals may be naive to threats (e.g. alien predators), unable to respond to threats (e.g. scarce prey), or the threat may be too minimal to warrant a response. Some environments more easily facilitate avoidance (e.g. high food and shelter accessibility), and so accounting for landscape context is key. A threat response may be routine, or only employed or enhanced when the threat is heightened. This likely depends on how constant and predictable the threat is.

Animals may respond to threats by altering their activity patterns (i.e. avoidance). An animal may avoid locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’). Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness. Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.

@@

Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?). For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations. Using GPS data and habitat selection methods,  Kohl et al. 2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs. However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour. And so, will elk respond to wolves in the same way in different environmental contexts? There are little-no examples of replicated spatiotemporal interactions studies.

There is contention around the interactions between two invasive predators - feral cats and foxes. Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats. Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal. Here, we aim to dig into the behavioral response further - in particular diel activity times.

@@

## Camera-trap surveys

We compiled camera-trap data from three distinct studies across the two regions. After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites. Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights. Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study. Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019.

@@

## Generalised additive models 
We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017]. We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa]. We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e. the effect of unmodelled environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions. This formed the base model specification for each model we fitted. Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below). We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e. shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372]. We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].

We expected the activity of each predator species to follow a broadly consistent pattern (e.g. lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1). Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates. We therefore specified a global smoother for hour (i.e. average response), along with group-level smoothers with shared wiggliness for each EVC group [i.e. model GS in @pedersen2019hierarchical]. This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.

To investigate changes in feral cat  diel activity in response to counts of foxes. We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').

@@

We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type. By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions. This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.

@@

>>>>

---
title: Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
    footnote: 
  - name: Alan Robley
    email: 
    affiliation: ARI
  - name: Nick Golding
    email: 
    affiliation: CU,TKI
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
  - code: ARI 
    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
  - code: CU 
    address: Curtin University, Bentley, WA, Australia
  - code: TKI 
    address: Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  1. Background.

  2. Methods.

  3. Results.

  4. *Synthesis and applications.*

@@

Diel activity patterns - the distribution of activity throughout the daily cycle - are an often overlooked behavioural aspect in ecology. A species must allocate particular times of the day to rest, foraging, socialisation and reproduction; which are traded-off against threat avoidance and deterrence against resource loss. These trade-offs may or may not incur costs to individual and population fitness.Diel activity may be bound by physiological and environmental constraints. For example, animal must be nocturnal as to not overheat or because that is when prey is most accessible. Other animals have varied diel activity, which may alter their ecosystem impact. Rigid or flexible animal diel activity may be the difference between adapting or succumbing to a threat.

Animals may respond to threats by altering their activity patterns (i.e. avoidance). An animal may avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’).  Modelling approaches which do not consider both spatial and temporal niches in a combined framework miss important nuance in animal behaviour. This is particularly the case when threats vary spatiotemporally, such as other species.

@@

Dynamic behavioural responses are important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?). For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations. Using GPS data and habitat selection methods,  Kohl et al. 2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs. However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour. And so, will elk respond to wolves in the same way in different environmental contexts? There are little-no examples of replicated spatiotemporal interactions studies.

Animals may be naive to threats (e.g. alien predators), unable to respond to threats (e.g. scarce prey), or the threat may be too minimal to warrant a response. Some environments more easily facilitate avoidance (e.g. high food and shelter accessibility), and so accounting for landscape context is key. A threat response may be routine, or only employed or enhanced when the threat is heightened. This likely depends on how constant and predictable the threat is.

Invasive species are highly adaptable to new conditions, which may facilitated by flexible spatiotemporal behaviour. 
There is contention around the interactions between two invasive predators - feral cats and foxes. Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats. Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioural responses (ranging and detectability) were not universal. Here, we aim to dig into the behavioural response further - in particular diel activity times.

@@

## Camera-trap surveys
We compiled camera-trap data from three distinct studies across the two regions. After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites. Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights. Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study. Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019.

@@

## Generalised additive models 
We modelled predator counts (response variable) with generalised additive models (hereafter 'GAMs') implemented in the 'mgcv' R-package [@wood2017]. GAMs learn from the data to specify nonlinear relationships between response and explanatory variables. A maximum wiggliness is specified for each explanatory variable, with complexity automatically reduced (all the way to completely linear function) through a smoothing penalty to avoid overfitting. We used REML estimation and the negative binomial family (as overdispersion, but not zero-inflation, was initially detected with a Poisson distribution using the 'DHARMa' R-package [@DHARMa]).

We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e. the effect of unaccounted for environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions. This formed the base model specification for each model we fitted. Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore variations in (1) predator diel activity across EVC groups, (2) feral cat diel activity across the gradient of observed fox counts, as well as (3) individual feral cat diel activity, which are detailed in the sections below. We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].

### Variation in predator diel activity across vegetation types
We expected the activity of each predator species to follow a broadly consistent pattern (e.g. lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1). Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates. We therefore specified a global smoother for hour (i.e. average response), along with group-level smoothers with shared wiggliness for each EVC group [i.e. model GS in @pedersen2019hierarchical]. This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.

### Variation in feral cat diel activity across apex predator activity
To investigate changes in feral cat  diel activity in response to counts of foxes. We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').

@@

We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type. We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e. shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372]. This meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions. This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.

### Individual variation in feral cat diel activity

@@

GRAVEYARD
===================

Diel activity patterns are a product of physiological constraints, species adaptiveness and landscape context. These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time. This is key because species interactions and niche preferences are multidimensional.

Effective wildlife management relies on understanding species behavioural and population dynamics to accurately predict intervention outcomes. Behavioural dynamics can vary across individuals, environmental conditions and with the activity of different species, and this may or may not lead to population-level impacts.

Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g. fire).

### 2021-08-06 manuscript/manuscript.tex
update

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Background.
  \item
    Methods.
  \item
    Results.
  \item
    \emph{Synthesis and applications.}
  \end{enumerate}
  \end{abstract}
   \begin{keyword} daily activity patterns; camera trap; \emph{Felis catus}; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; \emph{Vulpes vulpes}.\end{keyword}
 \end{frontmatter}

@@

Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes. These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time. This is key because species interactions and niche preferences are multidimensional.

Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire). Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response. Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key. A threat response may be routine, or only employed or enhanced when the threat is heightened. This likely depends on how constant and predictable the threat is.

Animals may respond to threats by altering their activity patterns (i.e.~avoidance). An animal may avoid locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance'). Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness. Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.

@@

Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?). For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations. Using GPS data and habitat selection methods, Kohl et al.~2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs. However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour. And so, will elk respond to wolves in the same way in different environmental contexts? There are little-no examples of replicated spatiotemporal interactions studies.

There is contention around the interactions between two invasive predators - feral cats and foxes. Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats. Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal. Here, we aim to dig into the behavioral response further - in particular diel activity times.

@@

We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter `GAMMs') implemented in the `mgcv' R-package (Wood 2017). We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the `DHARMa' R-package (Hartig 2020). We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unmodelled environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions. This formed the base model specification for each model we fitted. Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below). We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011). We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).

@@

@@

We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type. By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions. This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.

@@

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    Background.
  \item
    Methods.
  \item
    Results.
  \item
    \emph{Synthesis and applications.}
  \end{enumerate}
  \end{abstract}
   \begin{keyword} daily activity patterns; camera trap; \emph{Felis catus}; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; \emph{Vulpes vulpes}.\end{keyword}
 \end{frontmatter}

@@

Diel activity patterns - the distribution of activity throughout the daily cycle - are an often overlooked behavioural aspect in ecology. A species must allocate particular times of the day to rest, foraging, socialisation and reproduction; which are traded-off against threat avoidance and deterrence against resource loss. These trade-offs may or may not incur costs to individual and population fitness.Diel activity may be bound by physiological and environmental constraints. For example, animal must be nocturnal as to not overheat or because that is when prey is most accessible. Other animals have varied diel activity, which may alter their ecosystem impact. Rigid or flexible animal diel activity may be the difference between adapting or succumbing to a threat.

Animals may respond to threats by altering their activity patterns (i.e.~avoidance). An animal may avoid times of the day when a threat is likely to occur (hereafter `temporal avoidance'), or locations where the threat is likely to occur (hereafter `spatial avoidance'), avoid or avoid both locations and times a threat is heightened (hereafter `spatiotemporal avoidance'). Modelling approaches which do not consider both spatial and temporal niches in a combined framework miss important nuance in animal behaviour. This is particularly the case when threats vary spatiotemporally, such as other species.

@@

Dynamic behavioural responses are important from a management perspective because they want to know whether an action will cause a change in animal behaviour / range?). For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations. Using GPS data and habitat selection methods, Kohl et al.~2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs. However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour. And so, will elk respond to wolves in the same way in different environmental contexts? There are little-no examples of replicated spatiotemporal interactions studies.

Animals may be naive to threats (e.g.~alien predators), unable to respond to threats (e.g.~scarce prey), or the threat may be too minimal to warrant a response. Some environments more easily facilitate avoidance (e.g.~high food and shelter accessibility), and so accounting for landscape context is key. A threat response may be routine, or only employed or enhanced when the threat is heightened. This likely depends on how constant and predictable the threat is.

Invasive species are highly adaptable to new conditions, which may facilitated by flexible spatiotemporal behaviour.
There is contention around the interactions between two invasive predators - feral cats and foxes. Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats. Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioural responses (ranging and detectability) were not universal. Here, we aim to dig into the behavioural response further - in particular diel activity times.

@@

We modelled predator counts (response variable) with generalised additive models (hereafter `GAMs') implemented in the `mgcv' R-package (Wood 2017). GAMs learn from the data to specify nonlinear relationships between response and explanatory variables. A maximum wiggliness is specified for each explanatory variable, with complexity automatically reduced (all the way to completely linear function) through a smoothing penalty to avoid overfitting. We used REML estimation and the negative binomial family (as overdispersion, but not zero-inflation, was initially detected with a Poisson distribution using the `DHARMa' R-package (Hartig 2020)).

We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e.~the effect of unaccounted for environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions. This formed the base model specification for each model we fitted. Models (n = 3) differed in their specification of the cyclical hour smooth in order to explore variations in (1) predator diel activity across EVC groups, (2) feral cat diel activity across the gradient of observed fox counts, as well as (3) individual feral cat diel activity, which are detailed in the sections below. We plotted models using `ggplot2' and `gratia' R-packages (Wickham 2016, Simpson 2021).

\hypertarget{variation-in-predator-diel-activity-across-vegetation-types}{%
\subsubsection{Variation in predator diel activity across vegetation types}\label{variation-in-predator-diel-activity-across-vegetation-types}}

@@

\hypertarget{variation-in-feral-cat-diel-activity-across-apex-predator-activity}{%
\subsubsection{Variation in feral cat diel activity across apex predator activity}\label{variation-in-feral-cat-diel-activity-across-apex-predator-activity}}

@@

We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type. We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e.~shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model (Marra and Wood 2011). This meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions. This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.

\hypertarget{individual-variation-in-feral-cat-diel-activity}{%
\subsubsection{Individual variation in feral cat diel activity}\label{individual-variation-in-feral-cat-diel-activity}}

@@

\hypertarget{graveyard}{%
\section{GRAVEYARD}\label{graveyard}}

Diel activity patterns are a product of physiological constraints, species adaptiveness and landscape context. These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time. This is key because species interactions and niche preferences are multidimensional.

Effective wildlife management relies on understanding species behavioural and population dynamics to accurately predict intervention outcomes. Behavioural dynamics can vary across individuals, environmental conditions and with the activity of different species, and this may or may not lead to population-level impacts.

Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g.~fire).
